---
title: Parallel computation
author: Marc Choisy
date: '2018-05-30'
slug: parallel-computation
categories:
  - R
  - computing
tags: []
editor_options: 
  chunk_output_type: console
---


```{r include = FALSE}
# general options:
knitr::knit_hooks$set(margin = function(before, options, envir) {
  if (before) par(mgp = c(1.5, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
  else NULL
})

knitr::opts_chunk$set(margin = TRUE, prompt = TRUE, comment = "", eval = FALSE,
                      collapse = TRUE, cache = FALSE, autodep = TRUE,
                      dev.args = list(pointsize = 11), fig.height = 3.5,
                      fig.width = 4.24725, fig.retina = 2, fig.align = "center")

```

## Parallel version of `replicate` with `parallel`

Loading the package:

```{r}
library(parallel)
```

of which we'll use the 6 following functions:

* `detectCores` which detects the number of CPU cores;
* `makeCluster` which creates a cluster of a given number of cores;
* `clusterEvalQ` which evaluates a literal expression on each cluster node. It
is a parallel version of `evalq`.
* `clusterExport` which assigns the values on the master R process of the
variables named in its named list argument to variables of the same names in the
global environment (aka workspace) of each node;
* `parSapply` parallel version of `sapply`.
* `stopCluster` which shuts done the cluster.

The first step is to create a cluster. It's good practice to let one core
available, in case:

```{r}
cl <- makeCluster(detectCores() - 1)
```

Then, we need to populate each core of the cluster with any package we may need:

```{r}
clusterEvalQ(cl, library(package1, package2))
```

as well as any object (data and functions) we may need from the current
workspace, for example:

```{r}
clusterExport(cl, c("data1", "data2", "fct1", "fct2"))
```

Then, you ready to launch your calculations with `nb` replicates:

```{r}
out <- parSapply(cl, 1:nb, function(x) {
# here you include what should be done in each replication.
# this calculation will require the packages package1 and package2,
# as well as data data1 and data2 and function fct1 and fct2 of the
# workspace.
})
```

Once your calculations are done, you have to terminate the cluster:

```{r}
stopCluster(cl)
```

