---
title: Spatial buffers
author: Marc Choisy
date: '2018-01-22'
slug: spatial-buffers
categories:
  - R
  - spatial
  - statistics
tags: []
---

```{r include = FALSE}
# general options:
knitr::knit_hooks$set(margin = function(before, options, envir) {
  if (before) par(mgp = c(1.5, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
  else NULL
})

knitr::opts_chunk$set(margin = TRUE, prompt = TRUE, comment = "",
                      collapse = TRUE, cache = TRUE, autodep = TRUE,
                      dev.args = list(pointsize = 11), fig.height = 3.5,
                      fig.width = 4.24725, fig.retina = 2, fig.align = "center")

```

## Packages needed

Install these packages if they are not already installed on your system, then
load them:

```{r message = FALSE}
library(sp)        # plot, points, spsample
library(purrr)     # map2
library(dplyr)     # %>%, mutate
library(rgeos)     # gIntersection
library(raster)    # plot, buffer
library(geosphere) # areaPolygon
```

We can compute statistics in buffer zones around specific points. For that we
need spatial data in raster format. These data can be qualitative (e.g. land
cover) or quantitative (e.g elevation or population size).

## Defining buffers zone and their areas

Let's choose 10 points (cases) at random inside the polygon of Vietnam. First
the polygon of Vietnam:

```{r}
# devtools::install_github("choisy/gadmVN")
# devtools::install_github("choisy/sptools")
vn <- gadmVN::gadm(level = "country") %>% sptools::largest_polygons()
```

Let's see that polygon:

```{r}
plot(vn)
```

Let's now choose 10 random locations inside the polygon of Vietnam:

```{r}
set.seed(30101976) # for the purpose of the reproducibility here only
nb <- 10
cases <- spsample(vn, nb, "random")
```

Let's see that:

```{r}
plot(vn)
points(cases, pch = 3, col = "red")
```

Let's now define buffer zones around these points with a radius of 20 km:

```{r}
radius <- 20000 # in meters
buffers <- buffer(cases, radius, dissolve = FALSE)
```

Let's see that:

```{r}
plot(vn, col = adjustcolor("blue", .05))
plot(buffers, col = adjustcolor("red", .1), add = TRUE)
```

Note that the 8th of these buffers is crossed the border of Vietnam:

```{r}
plot(buffers[8], col = adjustcolor("red", .05))
plot(vn, col = adjustcolor("blue", .1), add = TRUE)
box(bty = "o")
```

For that reason, we first need to compute the intersection of the buffers with
the polygon of Vietnam before we can compute the areas of these buffers:

```{r}
(areas <- sapply(seq_along(buffers),
                 function(x) areaPolygon(gIntersection(vn, buffers[x])) / 10^6))
```

Note that the area of the 8th buffer is slightly less that the areas of the
other buffers.

## Quantitative statistics

Let's start we a quantitative example, looking at population size  Vietnam:

```{r}
# devtools::install_github("choisy/worldpopVN")
wpvn <- worldpopVN::getpop(2015)
```

The population size for Vietnam looks like this:

```{r}
plot(wpvn)
```

Let's extract the values inside each buffers and sum them:

```{r}
(popsizes <- raster::extract(wpvn, cases, buffer = radius) %>% sapply(sum, na.rm = TRUE))
```

To get the population density (/ km2) in this buffer zones, we just need to
divide this values by the areas:

```{r}
(densities <- popsizes / areas)
```

### A pipeline version

Let's describe here a pipeline version of the code above. Let's imagine that we
start from a matrix of coordinates:

```{r}
(coord_mat <- coordinates(cases))
```

The pipeline below used the above-defined objects `radius`, `vn` and `wpvn`, and
also the projection `llCRS` defined below:

```{r}
llCRS <- CRS("+proj=longlat +datum=WGS84 +no_defs +ellps=WGS84 +towgs84=0,0,0")
```

Let's now then define 2 utilitary functions:

```{r}
calc_area <- function(sp, radius, polygon) {
  sp %>%
    buffer(radius) %>%
    gIntersection(polygon) %>%
    areaPolygon()
}
```

and

```{r}
calc_popsize <- function(sp, radius, raster) {
  raster::extract(raster, sp, buffer = radius) %>% sapply(sum, na.rm = TRUE)
}
```

Then, the pipeline goes:

```{r}
coord_mat %>%
  data.frame %>% 
  mutate(sp      = map2(x, y, data.frame) %>% lapply(. %>% SpatialPoints(llCRS)),
         area    = sapply(sp, calc_area, radius, vn),
         popsize = sapply(sp, calc_popsize, radius, wpvn),
         density = popsize / area) %>%
  dplyr::select(-sp)
```


### A function version

We can put the above pipe into a function:

```{r}
buffering <- function(coord, radius, rast, polyg) {
  require(sp)
  require(purrr)
  require(rgeos)
  require(dplyr)
  require(raster)
  require(geosphere)
  stopifnot(proj4string(rast) == proj4string(polyg))
  coord %>%
  data.frame %>% 
      mutate(sp      = map2(x, y, data.frame) %>% lapply(. %>% SpatialPoints(llCRS)),
         area    = sapply(sp, calc_area, radius, vn),
         popsize = sapply(sp, calc_popsize, radius, wpvn),
         density = popsize / area) %>%
  dplyr::select(-sp)
}
```

Using that function:

```{r}
buffering(coord_mat, radius, wpvn, vn)
```

## Qualitative statistics

Let's see now how it works with qualitative variables such as land cover.

```{r}
gcvn <- globcoverVN::getgcvn()
```

The plot of the land cover with the buffers:

```{r}
plot(gcvn)
plot(buffers, add = TRUE)
```

To see the legend:

```{r}
globcoverVN::show_legend(gcvn)
```

Magnifying the 8th buffer:

```{r}
plot(buffers[8])
plot(gcvn, axes = FALSE, legend = FALSE, add = TRUE)
plot(buffers[8], add = TRUE)
plot(vn, add = TRUE)
box(bty = "o")
```

We can calculate the percentage of land cover for each buffer by using this
function:

```{r}
calc_globcov <- function(sp, radius, raster) {
  raster::extract(raster, sp, buffer = radius) %>% 
    table() %>% 
    (function(x) lapply(list(I, sum), function(f) f(x))) %>%
    do.call(`/`, .)
}
```

like this:

```{r}
coord_mat %>% 
  data.frame() %$% 
  map2(x, y, data.frame) %>%
  lapply(. %>% SpatialPoints(llCRS)) %>% 
  lapply(calc_globcov, radius, gcvn) %>% 
  do.call(dplyr::bind_rows, .) %>% 
  mutate_all(funs(ifelse(is.na(.), 0, .))) %>%
  bind_cols(data.frame(coord_mat), .)
```

