---
title: Computing elevation by province
author: Marc Choisy
date: '2019-10-23'
slug: computing-elevation-by-province
categories:
  - R
  - spatial
tags: []
editor_options: 
  chunk_output_type: console
---

```{r general options, include = FALSE}
knitr::knit_hooks$set(
  margin = function(before, options, envir) {
    if (before) par(mgp = c(1.5, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
    else NULL
  })

knitr::opts_chunk$set(cache = TRUE, autodep = TRUE, message = FALSE, warning = FALSE,
                      dev.args = list(pointsize = 11), fig.height = 3.5,
                      fig.width = 4.24725, fig.retina = 2, fig.align = "center")

# options(width = 137)
```

## Introduction

Here we show how to compute 2 different measures of elevation by province. The
first one is the simple mean of the elevation measures inside the polygon of the
province. The second one weights the mean by the local population density. For
that, we use the polygons of the country and the provinces from
[GADM](https://gadm.org), the SRTM raster elevation data from
[CGIAR](http://srtm.csi.cgiar.org) and raster population density data from
[WorldPop](https://www.worldpop.org).

## Packages

```{r}
library(sp)
library(raster)
```

## Simple mean

Let's downlaod the elevation data (751.5 MB) from
[SRTM](http://srtm.csi.cgiar.org):

```{r eval = FALSE, include = FALSE}
download.file("http://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_30x30/TIFF/N00E090.zip", "content/post/srtm.zip")
```

```{r eval = FALSE}
download.file("http://srtm.csi.cgiar.org/wp-content/uploads/files/srtm_30x30/TIFF/N00E090.zip", "srtm.zip")
```

Let's unzip it (making a file of 2.4 GB):

```{r eval = FALSE, include = FALSE}
unzip("content/post/srtm.zip", exdir = "content/post")
```

```{r eval = FALSE}
unzip("srtm.zip")
```

And load it into R:

```{r eval = FALSE, include = FALSE}
srtm <- raster("content/post/cut_n00e090.tif")
```

```{r}
srtm <- raster("cut_n00e090.tif")
```

which gives

```{r}
srtm
```

Note that the CRS is not defined... However, from 
[here](https://cgiarcsi.community/data/srtm-90m-digital-elevation-database-v4-1)
we can see that *"The data is projected in a Geographic (Lat/Long) projection,
with the WGS84 horizontal datum and the EGM96 vertical datum."* Let's thus
define this projection:

```{r}
crs(srtm) <- CRS("+init=EPSG:4326")
```

Let's check:

```{r}
srtm
```

We're good! This file is quite big and we need the data only for Vietnam. We
thus want to crop it. For that we need the polygon of Vietnam that we can
download from [GADM](https://gadm.org):

```{r eval = FALSE, include = FALSE}
download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_VNM_0_sp.rds", "content/post/country.rds")
```

```{r eval = FALSE}
download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_VNM_0_sp.rds", "country.rds")
```

Lets load it:

```{r eval = FALSE, include = FALSE}
country <- readRDS("content/post/country.rds")
```

```{r}
country <- readRDS("country.rds")
```

Which gives:

```{r}
country
```

We can see that this time the CRS is correctly defined and that it is the same
projection as for the SRTM raster file. Let's plot the elevation data together
with the country polygon:

```{r}
plot(srtm)
plot(country, add = TRUE)
```

Let's now use the country polygon to crop the raster file. Note that this can be
done only because the 2 objects have the same projection. If it were not the
case, we would need to reproject one of them according to the projection of the
other, using the `sp::spTranform()` function. It takes about 15''.

```{r}
srtm_cropped <- crop(srtm, country)
```

Let's write the output to disk:

```{r eval = FALSE, include = FALSE}
writeRaster(srtm_cropped, "content/post/srtm_cropped.tif")
```

```{r include = FALSE}
writeRaster(srtm_cropped, "srtm_cropped.tif", overwrite = TRUE)
```

```{r eval = FALSE}
writeRaster(srtm_cropped, "srtm_cropped.tif")
```

Let's reload it:

```{r eval = FALSE, include = FALSE}
srtm_cropped <- raster("content/post/srtm_cropped.tif")
```

```{r}
srtm_cropped <- raster("srtm_cropped.tif")
```

Let's see the result:

```{r}
plot(srtm_cropped)
plot(country, add = TRUE)
```

We can see that the cropped_version of the GeoTiff object is reduced from 2.4 GB
down to 194.6 MB:

```{r}
file.size(filename(srtm)) / 2^30
```

and:

```{r}
file.size(filename(srtm_cropped)) / 2^20
```

Noticing that in this rectangular, we acutally need only a small portion of the
pixels, we can go a step further in reducing the data by applying a mask. For
that, we first need to create a mask by basically rasterizing the polygon using
the same grid as the raster object. Note however that this step takes about 26', 
essentially because the size of the polygon is big. Given that, we'll see at the
end of this section that this step is actually not necessary.

```{r eval = FALSE}
country_mask <- rasterize(country, srtm_cropped)
```

```{r eval = FALSE, include = FALSE}
writeRaster(country_mask, "content/post/country_mask.tif")
```

```{r include = FALSE}
country_mask <- raster("country_mask.tif")
```

Now that the mask is done, we can use it on the cropped raster:

```{r}
srtm_masked <- mask(srtm_cropped, country_mask)
```

Let's save to disk and reload:

```{r eval = FALSE, include = FALSE}
writeRaster(srtm_masked, "content/post/srtm_masked.tif")
srtm_masked <- raster("content/post/srtm_masked.tif")
```

```{r include = FALSE}
writeRaster(srtm_masked, "srtm_masked.tif", overwrite = TRUE)
srtm_masked <- raster("srtm_masked.tif")
```

```{r eval = FALSE}
writeRaster(srtm_masked, "srtm_masked.tif")
srtm_masked <- raster("srtm_masked.tif")
```

We can see that masking allows to reduce the file size from 195 MB to 86 MB:

```{r}
file.size(filename(srtm_masked)) / 2^20
```

Let's see what it looks like:

```{r}
plot(srtm_masked)
plot(country, add = TRUE)
```

Great! Next step now, in order to compute a value of elevation per province, is
to get the polygons of the provinces:

```{r eval = FALSE, include = FALSE}
download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_VNM_1_sp.rds", "content/post/provinces.rds")
```

```{r eval = FALSE}
download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_VNM_1_sp.rds", "provinces.rds")
```

Let's load it:

```{r eval = FALSE, include = FALSE}
provinces <- readRDS("content/post/provinces.rds")
```

```{r}
provinces <- readRDS("provinces.rds")
```

Which gives:

```{r}
provinces
```

We can again verify that the projection is the same as the raster and the
country polygon. Let's see what it looks like:

```{r}
plot(provinces)
```

What we want to do now is, for each province, extracting the elevation data from
the raster file that are inside the polygon of the province and average them.
This is what the following function does:

```{r}
mean_ele <- function(x) {
  require(magrittr)
  prov <- provinces[x, ]
  rast <- crop(srtm_masked, prov) # cropping before the mask
  mask(rast, rasterize(prov, rast)) %>% 
    values() %>% 
    mean(na.rm = TRUE)
}
```

Note interestingly that a crop before the mask speeds up the masking by 6.
Interestingly too, this `mean_ele()` function is about twice as fast as the use
of the `extract()` function. Compare for example

```{r}
system.time(opt1 <- sapply(1:3, mean_ele))
```

with 

```{r}
system.time(opt2 <- sapply(extract(srtm_masked, provinces[1:3, ]), mean, na.rm = TRUE))
```

The reason for that is not clear to me. Maybe there are a number of checks in
`extract()` function that substantially slow everything down. Let's check that
it provides the same results however:

```{r}
opt1 - opt2
```

Let's also see what is the benefit of working on `strm_masked` instead of
`strm_cropped`:

```{r}
mean_ele2 <- function(x) {
  require(magrittr)
  prov <- provinces[x, ]
  rast <- crop(srtm_cropped, prov)
  mask(rast, rasterize(prov, rast)) %>% 
    values() %>% 
    mean(na.rm = TRUE)
}
system.time(opt3 <- lapply(1:3, mean_ele2))
```

So, it means that the only advantage of the `srtm_masked` over `srtm_cropped` is
that it's about twice as light on the disk but it doesn't make the calculations
based on these rasters any faster. Good to know! Let's now compute this for all
the provinces, which takes about 10':

```{r}
elevations1 <- sapply(seq_along(provinces), mean_ele)
```

## Weighting by population densities

Now, let's consider an alternative way of computing these elevations by province,
weighting them by the local population density. We can retrieve the local
population density for Vietnam as a raster file from
[WorldPop](http://www.worldpop.org):

```{r eval = FALSE, include = FALSE}
download.file("ftp://ftp.worldpop.org.uk/GIS/Population/Global_2000_2020/2010/VNM/vnm_ppp_2010.tif", "content/post/worldpop.tif")
```

```{r eval = FALSE}
download.file("ftp://ftp.worldpop.org.uk/GIS/Population/Global_2000_2020/2010/VNM/vnm_ppp_2010.tif", "worldpop.tif")
```

Let's load it:

```{r eval = FALSE, include = FALSE}
worldpop <- raster("content/post/worldpop.tif")
```

```{r}
worldpop <- raster("worldpop.tif")
```

Good news here is that the raster is already masked:

```{r}
plot(worldpop)
```

It also has the same projection than the other objects we've been dealing with
so far:

```{r}
worldpop
```

Of note too, it has the same resolution as `srtm_masked`, but not quite the same
grid definition though:

```{r}
srtm_masked
```

In order to be able to weigth the elevation data from `srtm_masked` by the
population density data from `worldpop` we need to ensure that the two rasters
use the same grid. This implies resampling one of the raster on the grid of the
other one. Let's resample `srtm_masked` on `worldpop`. It takes about 2':

```{r}
srtm_masked2 <- resample(srtm_masked, worldpop)
```

We can check that it worked:

```{r}
worldpop
```

and:

```{r}
srtm_masked2
```

Below is a new version of the `mean_ele()` function that performs the population
weighting:

```{r}
mean_ele3 <- function(x) {
  prov <- provinces[x, ]
  srtm <- crop(srtm_masked2, prov)
  wpop <- crop(worldpop, prov)
  prov_mask <- rasterize(prov, srtm)
  srtm_val <- values(mask(srtm, prov_mask))
  wpop_val <- values(mask(wpop, prov_mask))
  weights <- wpop_val / sum(wpop_val, na.rm = TRUE)
  sum(weights * srtm_val, na.rm = TRUE)
}
```

Let's apply it to all the provinces, it takes about 16':

```{r}
elevations2 <- sapply(seq_along(provinces), mean_ele3)
```

Let's compare the 2 measures of elevation by province:

```{r}
(elevations <- data.frame(province = provinces$VARNAME_1, elevations1, elevations2))
```

Let's visualize this:

```{r}
plot(elevations2 ~ elevations1, elevations, xlab = "weighted by population density", ylab = "non weighted", col = "blue")
abline(0, 1)
```

This figure shows clearly that the population tends to gather in the lower parts
of the province.
