---
title: Writing exercises in R
author: Marc Choisy
date: '2019-02-18'
slug: writing-exercises-in-r
categories:
  - R
  - Teaching
tags: []
editor_options: 
  chunk_output_type: console
---

The **literate programming** ability of Rmarkdown makes it a great language for
**reproducible research**. It is also used for many other applications (such as
writing books, websites, blogs, apps, etc...). One of them that I've found
particularly useful is for teaching, when the tutor needs to write exercises of
which (s)he wants to control the visibility of the solutions in a way or another.

Here I list 3 options that I've found particularly useful. The first one makes
use of the `include` **chunk option** of the Rmarkdown file and allows the tutor
to control whether the solutions will be visible or not (the options have to be
set before the compilation of the Rmarkdown document into HTML). The second
option makes use of the `code_folding` option of the **YAML header** of the
Rmarkdown document and leaves the control of the solutions visibility to the
student. By default, all the R code will be either visible or hidden (depending
on the value of the `code_folding` option in the YAML header), but all the R
code will necessarily have the same default option. If one is interested in
having some R code that is always visible and some other chunks of R code that
is hidden but could be shown by the user, then the third option making use of 
**JavaScript** would do the job. The insertion of JavaScript code in the
Rmarkdown file is the most flexible solution that allows to hide any number of
lines (not necessarily R chunks). However, compared to the 2 previous options,
in case an R code chunk is hidden, it will also hide the output of this R code
chunk (whereas it remains visible for the first 2 options). The table below
recaps the different characteristics of the 3 options.

  Method         |   Control   | Default   |   Which         |   Hidden output 
---------------- | ----------- | --------- | --------------- | -----------------
  chunk option   |   tutor     | NA        |   any R chunk   |   no
  YAML header    |   tutor     | any       |   all R chunk   |   no
  JavaScript     |   student   | hidden    |   any line      |   yes

The way I work as a tutor is that I write the Rmarkdown document, then compile
it to an HTML document that I host on a web server (for example 
[RPubs](http://rpubs.com)), of which I provide the URL to the student. Links to
examples and templates are also provided below.

## Option 1: the tutor controls the visibility of the solutions

When I, as the tutor, want to keep full control of the visibility of the 
solution, I do it thanks to the chunk option `include` of the R code chunks of
the Rmarkdown document. For example, the following Rmarkdown code:

````markdown
**Exercise 1:** Define `x` and `y` vectors as below
`r ''````{r exercise_1}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

**Solution to Exercise 1:**
`r ''````{r solution_1, include = FALSE}
plot(x, y)
```
````

will produce the following HTML output where the solution is not visible:

**Exercise 1:** Define `x` and `y` vectors as below
```{r}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

**Solution to Exercise 1:**
```{r solution_1, include = FALSE}
plot(x, y)
```

Now, if I decide to show the solution, I just have to switch the `include`
chunck option from `FALSE` to `TRUE`:

````markdown
**Exercise 1:** Define `x` and `y` vectors as below
`r ''````{r exercise_1}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

**Solution to Exercise 1:**
`r ''````{r solution_1, include = TRUE}
plot(x, y)
```
````

which produces the following HTML output after recompilation:

**Exercise 1:** Define `x` and `y` vectors as below
```{r exercise_1}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

**Solution to Exercise 1:**
```{r include = TRUE}
plot(x, y)
```

Et voilà!

### Showing several solutions at once

You may have several exercises, each with its solution. If you decide to show
several solutions at once, all you have to do is define a boolean variable and
use the value of this variable for the value of the `include` option of all the
chuncks you want to control the same way. Good practice is to define the boolean 
variable in a chunk at the beginning of the Rmarkdown document that contains all
the settings of the document. Here is an example:

````markdown
`r ''````{r settings, include = FALSE}
switch1 <- FALSE
```

**Exercise 1:** Define `x` and `y` vectors as below
`r ''````{r exercise_1}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

**Solution to Exercise 1:**
`r ''````{r solution_1, include = switch1}
plot(x, y)
```

**Exercise 2:** Make a linear model `y` as a function of `x`.

**Solution to Exercise 2:**
`r ''````{r solution_2, include = switch1}
model <- lm(y ~ x)
```

**Exercise 3:** Make a plot of `y` as a function of `x`, together with the 
linear model

**Solution to Exercise 3:**
`r ''````{r solution_3, include = FALSE}
plot(x, y)
abline(model)
```
````

That produces the following HTML output:


```{r settings, include = FALSE}
switch1 <- FALSE
```

**Exercise 1:** Define `x` and `y` vectors as below
```{r}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

**Solution to Exercise 1:**
```{r, include = switch1}
plot(x, y)
```

**Exercise 2:** Make a linear model `y` as a function of `x`.

**Solution to Exercise 2:**
```{r solution_2, include = switch1}
model <- lm(y ~ x)
```

**Exercise 3:** Make a plot of `y` as a function of `x`, together 
with the linear model

**Solution to Exercise 3:**
```{r solution_3, include = FALSE}
plot(x, y)
abline(model)
```

where none of the solutions is displayed. If you want to show 
solutions to exercises 1 and 2 all at once, all you have to do is 
change the value of the `switch1` boolean variable in the settings
chunk as so:

````markdown
`r ''````{r settings, include = FALSE}
switch1 <- TRUE
```
````

Now, when you recompile the document, the HTML output looks like:

```{r include = FALSE}
switch1 <- TRUE
```

**Exercise 1:** Define `x` and `y` vectors as below
```{r}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

**Solution to Exercise 1:**
```{r, include = switch1}
plot(x, y)
```

**Exercise 2:** Make a linear model `y` as a function of `x`.

**Solution to Exercise 2:**
```{r include = switch1}
model <- lm(y ~ x)
```

**Exercise 3:** Make a plot of `y` as a function of `x`, together 
with the linear model

**Solution to Exercise 3:**
```{r include = FALSE}
plot(x, y)
abline(model)
```

where the solutions to both exercises 1 and 2 are displayed but not 
the solution to exercise 3.

## Option 2: the student can unhide the solution him(her)self

If seeing the solution of the exercises is let to the student's decision, then
you can opt for an interactive HTML document, using the `code_folding` option of
the YAML header of the Rmarkdown document, as so:

````markdown
---
title: "R exercises"
output:
  html_document:
    code_folding: hide
---
````

Note that here, we've even set the `code_folding` option to the `hide` value
which means that, by default, the R code of all the chunks will be hidden and
the user can decide to reveal all or each of them separately. If the
`code_folding` was set to `show` instead, then, by default, the R code of all
the chunks will have been visible and the user could have decided to hide them,
again, all of them at once, or each of them separately. An example of the first
option (`code_folding: hide`) is visible
[here](http://rpubs.com/choisy/code_folding_hide) (with source code
[here](http://bit.ly/codefoldinghide)) and an example of the second option
(`code_folding: show`) is visible
[here](http://rpubs.com/choisy/code_folding_show) (with source code
[here](http://bit.ly/codefoldingshow)). Note by the way that the output of the
R code chunks are never hidden.

## Option 3: flexibility with JavaScript

Because option 2 makes use of an option of the YAML header of the Rmarkdown
document, it is impossible to define different behaviors on different R code
chunks. As of February 2019, the only way to the combine user control of option
2 with the by-chunk definition of option 1 is to make use of JavaScript code in
the Rmarkdown document as suggested
[here](https://stackoverflow.com/questions/37755037/how-to-add-code-folding-to-output-chunks-in-rmarkdown-html-documents/37839683#37839683)
(it seems that an Rmarkdown solution to do so is under development at the moment).
What you have to do is

#### 1. include the following javascript code at the end of the Rmarkdown document:

```{js eval = FALSE}
<script>
  $(".toggle").click(function() {
    $(this).toggleClass("open");
  });
</script>
```

#### 2. write a CCS file that contains the following CSS code:

```{css eval = FALSE}
.toggle {
  height: 1.55em;
  overflow-y: hidden;
}
.toggle.open {
  height: auto;
}
```

This CCS code defines the style of the button. In this example, let's call
this file `styles.css` and make a link to it in the YAML header of the 
Rmarkdown document:

````markdown
---
title: "R exercises"
output:
  html_document:
    css: styles.css
---
````

#### 3. add buttons to your code wherever needed as so:

````markdown
**Exercise 1:** Define `x` and `y` vectors as below
```{r}
set.seed(123523458)
x <- 1:10
y <- sample(x)
```
and plot `y` as a function of `x`.

<div class="toggle"><button>Solution</button>
`r ''````{r}
plot(x, y)
```
</div><br>
````

The above example includes a button that, when clicked on, shows the R code
`plot(x, y)` together with its output. (Note that it is a slightly different
behaviour from what we had with the YAML `code_folding` option where only the R
code was hidden, not its output). See an example 
[here](http://rpubs.com/choisy/code_folding_js) and its source code
[here](http://bit.ly/codefoldingjs) for which the corresponding CSS file is
[here](http://bit.ly/foldingcodecss) and should be saved in a file named
`styles.css` in the save directory as the Rmarkdown file before compilation.
Note also that the buttons could braket any lines, these lines containing chunks 
or simply text, or actually anything.
