---
title: Lengths of latitude and longitude degrees
author: Marc Choisy
date: '2019-03-06'
slug: lengths-of-latitude-and-longitude-degrees
categories:
  - R
  - spatial
tags: []
editor_options: 
  chunk_output_type: console
---

In geography or spatial analyses, it is common that lengths are expressed in
degree (minutes, seconds) instead of meter or km as we are more used to.
Because of the spheroidal shape of the earth, the length in km of one degree of
latitude or longitude depends on the latitude of the location. Formulas to do
the conversion are available [here](https://gis.stackexchange.com/questions/75528/understanding-terms-in-length-of-degree-formula) that we use below.

The length of 1 degree of latitude in km, in function of the latitude `x` in
decimal degree is given by the following function:

```{r}
latitude_length <- function(x) {
  x <- pi * x / 180
  (111132.92 - 559.82 * cos(2 * x) + 1.175 * cos(4 * x) - .0023 * cos(6 * x)) / 1000
}
```

let's generate a vector of latitude values between the equator and the north
pole:

```{r}
latitudes <- seq(0, 90, le = 512)
```

We can verify that the length of 1 degree of latitude varies very little between
the equator and the north pole:

```{r}
range(latitude_length(latitudes))
```

this is because the shape of the earth is very close to be spherical. Now, let's
look at the length in km of 1 degree of longitude as a function of the latitude
`x` in decimal degree, which is given by the following function:

```{r}
longitude_length <- function(x) {
  x <- pi * x / 180
  (111412.84 * cos(x) - 93.5 * cos(3 * x) + .118 * cos(5 * x)) / 1000
}
```

let's use this function to plot the length of 1 degree of longitude as a 
function of the latitude:

```{r}
plot(longitude_length(latitudes), latitudes, type = "l",
     xlab = "length of 1° of longitude (km)",
     ylab = "latitude (°)",
     xlim = c(0, 120),
     ylim = c(0, 100))
```

We can thus verify that a cell close to the equator is almost of square shape 
and becomes more and more triangular as we get closer to the north pole. To
finish, let's have a look at the variation of the length of 1 degree of 
longitude in the intertropical zone:

```{r}
tropics <- seq(0, 23.43682, le = 512)
range(longitude_length(tropics))
```

And in Vietnam:

```{r}
vietnam <- seq(8.563512, 23.392651, le = 512)
range(longitude_length(vietnam))
```





