---
title: Computing land cover by province
author: Marc Choisy
date: '2019-10-28'
slug: computing-land-cover-by-province
categories:
  - big data
  - computing
  - R
  - spatial
tags: []
editor_options: 
  chunk_output_type: console
---

```{r general options, include = FALSE}
knitr::knit_hooks$set(
  margin = function(before, options, envir) {
    if (before) par(mgp = c(1.5, .5, 0), bty = "n", plt = c(.105, .97, .13, .97))
    else NULL
  })

knitr::opts_chunk$set(cache = TRUE, autodep = TRUE, message = FALSE, warning = FALSE,
                      dev.args = list(pointsize = 11), fig.height = 3.5,
                      fig.width = 4.24725, fig.retina = 2, fig.align = "center")

# options(width = 137)
```

## Introduction



## Packages

```{r}
library(sp)
library(raster)
library(magrittr)
```

## Simple mean

```{r}
globcov <- globcoverVN::getgcvn()
```

which gives

```{r}
globcov
```

WGS84 unprojected data.

```{r}
plot(globcov)
```

Note that we can get access to the legend this way:

```{r}
globcov@legend
```

Next step now, in order to compute values of land cover per province, is to get
the polygons of the provinces from [GADM](https://gadm.org):

```{r eval = FALSE, include = FALSE}
download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_VNM_1_sp.rds", "content/post/provinces.rds")
```

```{r eval = FALSE}
download.file("https://biogeo.ucdavis.edu/data/gadm3.6/Rsp/gadm36_VNM_1_sp.rds", "provinces.rds")
```

Let's load it:

```{r eval = FALSE, include = FALSE}
provinces <- readRDS("content/post/provinces.rds")
```

```{r}
provinces <- readRDS("provinces.rds")
```

Which gives:

```{r}
provinces
```

Note that the projection is the same as the raster. Let's see what it looks
like:

```{r}
plot(globcov)
plot(provinces, add = TRUE)
```

What we want to do now is, for each province, extracting the land cover data
from the raster file that are inside the polygon of the province and then
compute proportions. This is what the following function does:

```{r}
proportion_lc <- function(x) {
  require(magrittr)
  prov <- provinces[x, ]
  rast <- crop(globcov, prov) # cropping before the mask
  mask(rast, rasterize(prov, rast)) %>% 
    values() %>% 
    na.exclude() %>% 
    table() %>% 
    {100 * . / sum(.)}
}
```

Note interestingly that, as noted [here](),

* a crop before the mask speeds up the masking by 6 and
* this `proportion_lc()` function is about twice as fast as the use of the
`extract()` function.

Let's now compute this for all the provinces, which takes about 2':

```{r}
land_cover1 <- provinces %>%
  seq_along() %>%
  lapply(proportion_lc) %>% 
  setNames(provinces@data$VARNAME_1) %>% 
  c(., list(.id = "province")) %>%
  do.call(dplyr::bind_rows, .) %>% 
  tidyr::replace_na(., as.list(setNames(rep(0, ncol(.)), names(.))))
```

## Weighting by population densities

Now, let's consider an alternative way of computing these propotions of land
cover by province, weighting them by the local population density. We can
retrieve the local population density for Vietnam as a raster file from
[WorldPop](http://www.worldpop.org):

```{r eval = FALSE, include = FALSE}
download.file("ftp://ftp.worldpop.org.uk/GIS/Population/Global_2000_2020/2010/VNM/vnm_ppp_2010.tif", "content/post/worldpop.tif")
```

```{r eval = FALSE}
download.file("ftp://ftp.worldpop.org.uk/GIS/Population/Global_2000_2020/2010/VNM/vnm_ppp_2010.tif", "worldpop.tif")
```

Let's load it:

```{r eval = FALSE, include = FALSE}
worldpop <- raster("content/post/worldpop.tif")
```

```{r}
worldpop <- raster("worldpop.tif")
```

It gives:

```{r}
plot(worldpop)
plot(provinces, add = TRUE)
```

It also has the same projection than the other objects we've been dealing with
so far:

```{r}
worldpop
```

However, it does not have the same resolution as `globcov`: there are about 10
times more pixels in `worldpop` than in `globcov`. Let's thus resample
`worldpop` on `globcov` (takes about 2'):

```{r}
worldpop2 <- resample(worldpop, globcov)
```

We can check that it worked:

```{r}
worldpop2
```

and:

```{r}
globcov
```

Below is a new version of the `proportion_lc()` function that performs the
population weighting:

```{r}
proportion_lc2 <- function(x) {
  require(magrittr)
  prov <- provinces[x, ]
  glcv <- crop(globcov, prov)
  wpop <- crop(worldpop2, prov)
  prov_mask <- rasterize(prov, glcv)
  glvc_val <- values(mask(glcv, prov_mask))
  wpop_val <- values(mask(wpop, prov_mask))
  weights <- wpop_val / sum(wpop_val, na.rm = TRUE)
  data.frame(glvc_val, weights) %>% 
    na.exclude() %>%
    dplyr::group_by(glvc_val) %>%
    dplyr::summarize(val = sum(weights)) %$%
    setNames(val, glvc_val) %>%
    {100 * . / sum(.)}
}
```

Let's apply it to all the provinces, it takes about 2':

```{r}
land_cover2 <- provinces %>%
  seq_along() %>%
  lapply(proportion_lc2) %>% 
  setNames(provinces@data$VARNAME_1) %>% 
  c(., list(.id = "province")) %>%
  do.call(dplyr::bind_rows, .) %>% 
  tidyr::replace_na(., as.list(setNames(rep(0, ncol(.)), names(.))))
```

Let's compare the 2 measures of land cover by province:

```{r}
for (i in 2:ncol(land_cover1)) {
  x <- unlist(land_cover1[, i])
  y <- unlist(land_cover2[, i])
  xylim <- range(c(x, y))
  plot(y, x, xlim = xylim, ylim = xylim,
       xlab = "weighted by population density",
       ylab = "non weighted", col = "blue",
       main = globcov@legend@names[i - 1], cex.main = .5)
  abline(0, 1)
}
```

This figure shows clearly that the population tends to gather in the lower parts
of the province.
