---
title: The power of bash
author: Marc Choisy
date: '2019-06-07'
slug: the-power-of-bash
categories:
  - bash
tags: []
editor_options: 
  chunk_output_type: console
---

Very often I remember the name of the R function I need to use for a particular
problem but I don't quite remember how to use it exactly... And, even worse,
what I remember is that last time I had to use this function it took me a long
time to figure out how to use it... Since I'm not ready to go through that
learning process again, I decide to look for the R script on my computer that
did make use of this function. And here comes the power of `bash`! (if you have
bash on your computer, of course...). Well, to be more precise, I actually
refering to the particular `find` and `grep` commands here that can be used in
`bash` of course (if you're on linux or macos), but also in any other UNIX-type
shell. Let's demonstrate its power, using my specific example.

The R function I wanted to use was `sf::st_simplify()`. I knew I used it
recently (in the last 2 months) in an Rmarkdown file, but I couldn't remember
which one. Looking for all the Rmarkdown files in the current file hierarchy is
done by

```{r engine = "bash", eval = FALSE}
find ~/Dropbox/aaa -name "*.Rmd" -mtime -10w -type f
```

The `find` command is followed by the path of the location where we want to
start the file search (here `~/Dropbox/aaa`). Then it is followed by a number of
options specified by pairs of options names (preceded by the `-` sign) followed
by the option value. For example, here `-name "*.Rmd"` specifies that we are
looking for files with names ending with the `.Rmd` extension, `-mtime -10w`
means that we are looking at files that were modified less than 10 weeks ago and
`-type f` means that we are looking at regular files (excluding directories,
links, etc...). If we had been interested in file that were modified more than
3 months ago instead of less than 10 weeks ago, that would have been specified
by `-mtime +3m`. Among these options, one is particularly amazing in the sense
that it allows to run another command on the files that were selected. This
option is `-exec` and is followed by the command we want to run. For example,
here, I want to search for the `st_simplify` pattern in the selected files:

```{r engine = "bash", eval = FALSE}
find ~/Dropbox/aaa -name "*.Rmd" -mtime -10w -type f -exec grep -n "st_simplify" {} \;
```

Here it says that, on each of the files that were selected by the first part of
the `find` command (this files are here specified by the `{}` placeholder), we
want to execute the `grep` command (with the -n option in order to also print
the line number where the pattern was found in the file). Don't forget here the
`\;` sign that closes the value of the `-exec` option. Indeed, contrary to the
other options of the `find` command that have, as a value, only one "word" (i.e.
character string without any space), the `-exec` may have (and typically have)
many words as its value. Thus, the value of the `-exec` command is all the words
that are between `exec` and `\;`. Adding the `-print` option (no value) will have
the effect to print the name of the file where the pattern was found:

```{r engine = "bash", eval = FALSE}
find ~/Dropbox/aaa -name "*.Rmd" -mtime -10w -type f -exec grep -n "st_simplify" {} \; -print
```

And here it is!
